---
title: "Getting started with GeoThinneR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with GeoThinneR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
run_code <- requireNamespace("sf", quietly = TRUE) &
  requireNamespace("terra", quietly = TRUE) &
  requireNamespace("ggplot2", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = run_code
)
```

# Overview

**GeoThinneR** is an R package designed for efficient spatial thinning of point data, particularly useful for species occurrence data and other spatial analyses. It implements optimized neighbor search algorithms based on *kd*-trees, providing an **efficient, scalable, and user-friendly** approach. When working with large-scale datasets containing hundreds of thousands of points, applying thinning manually is not possible, and existing tools in R providing thinning methods for species distribution modeling often lack scalability and computational efficiency.

Spatial thinning is widely used across many disciplines to **reduce the density of spatial point data** by selectively removing points based on specified criteria, such as threshold spacing or grid cells. This technique is applied in fields like remote sensing, where it improves data visualization and processing, as well as in geostatistics, and spatial ecological modeling. Among these, species distribution modeling is a key area where thinning plays a crucial role in mitigating the effects of spatial sampling bias.

This vignette provides a step-by-step guide to thinning with **GeoThinneR** using both simulated and real-world datasets.

# Setup and loading data

To get started with **GeoThinneR**, we begin by loading the packages used in this vignette:

```{r setup, message=FALSE, warning=FALSE}
library(GeoThinneR)
library(terra)
library(sf)
library(ggplot2)
```

We'll use two dataset to demonstrate the use of **GeoThinneR**:
1. A **simulated a dataset** with `n = 2000` random occurrences randomly distributed within a 40 x 20 degree area for two artificial species.
2. A **real-world dataset** containing loggerhead sea turtle (*Caretta caretta*) occurrences in the Mediterranean Sea downloaded from GBIF (\url{https://www.gbif.org/}) and included with the package.

```{r load data, message=FALSE, results='hide'}
# Set seed for reproducibility
set.seed(123)

# Simulated dataset
n <- 2000
sim_data <- data.frame(
  lon = runif(n, min = -10, max = 10),
  lat = runif(n, min = -5, max = 5),
  sp = sample(c("sp1", "sp2"), n, replace = TRUE)
)

# Load real-world occurrence dataset
data("caretta")

# Load Mediterranean Sea polygon
medit <- system.file("extdata", "mediterranean_sea.gpkg", package = "GeoThinneR")
medit <- sf::st_read(medit)
```

Let's visualize the datasets before thinning:

```{r plot loaded data, fig.show="hold", out.width="50%", echo = FALSE}
ggplot() +
  geom_point(data = sim_data, aes(x = lon, y = lat, shape = sp), color = "#5183B3") +
  scale_shape_manual(values = c(sp1 = 16, sp2 = 17)) +
  xlab("Longitude") + ylab("Latitude") +
  ggtitle("Simulated Species Occurrences") +
  theme_minimal()

ggplot() +
  geom_sf(data = medit, color = "#353839", fill = "antiquewhite", alpha = 0.7)  +
  geom_point(data = caretta, aes(x = decimalLongitude, y = decimalLatitude), color = "#5183B3", alpha = 1) +
  xlab("Longitude") + ylab("Latitude") +
  ggtitle("C. caretta Mediterranean Sea Occurrences") +
  theme(
      panel.grid.major = element_line(color = gray(.5),linetype = "dashed",linewidth = 0.5),
      panel.background = element_rect(fill = "white"),
      axis.title = element_blank(),
      legend.position = "bottom"
    )
```

# Quick start guide

Now that our data is ready, let's use **GeoThinneR** for spatial thinning. All methods are accessed through the `thin_points()` function, which requires, at a minimum, a dataset containing georeferenced points (WGS84 CRS - EPSG:4326) in `matrix`, `data.frame`, or `tibble` format. The query dataset is provided through the `data` argument, and with `lon_col` and `lat_col` parameters, we can define the column names containing longitude and latitude (or x, y) coordinates. Default values are "lon" and "lat", and if not present, it will take the first two columns by default.

The **thinning method** is defined using the `method` parameter, with the following options:

- `"distance"`: Distance-based thinning  
- `"grid"`: Grid-based thinning  
- `"precision"`: Precision-based thinning  

The **thinning constraint** (distance, grid resolution or decimal precision) depends on the method chosen (these will be detailed in the next section). Regardless of the method, the algorithm selects which points to retain and remove. Since this selection is non-deterministic, you can run **multiple randomized thinning trials** to maximize the number of retained points.

Let’s quickly thin the simulated dataset using a simple distance-based method:

```{r}
# Apply spatial thinning to the simulated data
quick_thin <- thin_points(
  data = sim_data,     # Dataframe with coordinates
  lon_col = "lon",     # Longitude column name
  lat_col = "lat",     # Latitude column name
  method = "distance", # Method for thinning
  thin_dist = 20,      # Thinning distance in km,
  trials = 5,          # Number of trials
  all_trials = TRUE,   # Return all trials
  seed = 123           # Seed for reproducibility
)

quick_thin
```

This command applies spatial thinning ensuring no two retained points are closer than 20 km (`thin_dist`). The algorithm runs 5 random trials (`trials`) and return the results of all trials (`all_trials = TRUE`).

### Exploring the GeoThinned object

The output of `thin_points()` is a `GeoThinned` object, which contains:
- `x$retained`: a list of logical vectors indicating which points were retained in each trial
- `x$method`: the method used for thinning
- `x$params`: the parameters used for thinning
- `x$original_data`: the original dataset

You can apply basic methods for inspecting the object such as `print()`, `summary()`, and `plot()`. The `summary()` function provides detailed information about the thinning process, including the number of points retained in each trial, the nearest neighbor distance (NND), and the coverage of the retained points.

```{r}
summary(quick_thin)
```

We see that from 2000, we retained the 69.6% of points (1391), and the minimum distance between points has increased without reducing too much the spatial coverage or extent of the points compared to the original dataset.

```{r}
plot(quick_thin)
```


If we check the number of points retained in each trial, we can see that the 5 trials return different numbers of points retained. This is due to the randomness of the algorithm when selecting which points to retain. You can use `seed` for reproducibility. By setting, `all_trials = FALSE` it will just return the trial that kept the most points.

```{r}
# Number of kept points in each trial
sapply(quick_thin$retained, sum)
```

You can access the thinned dataset that retained the maximum number of points using the `largest()` function.

```{r}
head(largest(quick_thin))
```

If you want to access points from a specific trial, you can use the `get_trial()` function, which returns the thinned dataset for that trial.

```{r}
head(get_trial(quick_thin, trial = 2)) 
```

Just in case you want to get the points that were removed from the original dataset, you can use the `retained` list to index the original data. The `retained` list contains logical vectors indicating which points were retained in each trial. You can use this to filter the original data and get the removed points.

```{r}
trial <- 2
head(quick_thin$original_data[!quick_thin$retained[[trial]], ])
```

Also, you can use `as_sf()` to convert the thinned dataset to a `sf` object, which is useful for plotting and spatial analysis.

```{r}
sf_thinned <- as_sf(quick_thin)
class(sf_thinned)
```

### Real-world example

GeoThinneR includes a dataset of loggerhead sea turtle (*Caretta caretta*) occurrence records in the Mediterranean Sea, sourced from GBIF and filtered. We now apply a 30 km thinning distance, returning only the largest trial (the one with the maximum number of points retained, `all_trials` set to `FALSE`).

```{r}
# Apply spatial thinning to the real data
caretta_thin <- thin_points(
  data = caretta, # We will not specify lon_col, lat_col as they are in position 1 and 2
  lat_col = "decimalLatitude",
  lon_col = "decimalLongitude",
  method = "distance",
  thin_dist = 30, # Thinning distance in km,
  trials = 5,
  all_trials = FALSE,
  seed = 123
)

# Thinned dataframe stored in the first element of the retained list
identical(largest(caretta_thin), get_trial(caretta_thin, 1))
```
```{r, echo = FALSE}
ggplot() +
  geom_sf(data = medit, color = "#353839", fill = "antiquewhite", alpha = 0.7)  +
  geom_point(data = caretta, aes(x = decimalLongitude, y = decimalLatitude),color = "#EB714B", alpha = 0.5) +
  geom_point(data = largest(caretta_thin), aes(x = decimalLongitude, y = decimalLatitude),color = "#5183B3", alpha = 1) +
  xlab("Longitude") + ylab("Latitude") +
  ggtitle("C. caretta Mediterranean Sea Occurrences") +
  theme(
      panel.grid.major = element_line(color = gray(.5),linetype = "dashed",linewidth = 0.5),
      panel.background = element_rect(fill = "white"),
      axis.title = element_blank(),
      legend.position = "bottom"
    )
```

As you can see, with **GeoThinneR**, it's very easy to apply spatial thinning to your dataset. In the next section we explain the different thinning methods and how to use them, and also the implemented optimizations for large-scale datasets.

# Thinning methods

**GeoThinneR** offers different algorithms for spatial thinning. In this section, we will show how to use each thinning method, highlighting its unique parameters and features.

## Distance-based thinning

Distance-based thinning (`method="distance"`) ensures that all retained points are separated by at least a user-defined minimum distance `thin_dist` (in km). For geographic coordinates, using `distance = "haversine"` (default) computes great-circle distances, or alternatively, the Euclidean distance between 3D Cartesian coordinates (depending on the method) to identify neighboring points within the distance threshold. The algorithm iterates through the dataset and removes points that are too close to each other.

In general, the choice of search method depends on the size of the dataset and the desired thinning distance:

- `search_type="brute"`: Useful for small datasets, low scalability.
- `search_type="kd_tree"`: More memory-efficient, better scalability for medium-sized datasets.
- `search_type="local_kd_tree"` (default): Faster than `"kd_tree"` and reduces memory usage. Can be parallelized to improve speed.
- `search_type="k_estimation"`: Very fast when the number of points to remove is low or not very clustered, if not similar or worse than `"local_kd_tree"`.

You can find the benchmarking results for each method in the package manuscript (see references).

```{r}
dist_thin <- thin_points(sim_data, method = "distance", thin_dist = 25, trials = 3)
plot(dist_thin)
```


### Brute force

This is the most common greedy method for calculating the distance between points (`search_type="brute"`), as it directly computes all pairwise distances and retains points that are far enough apart based on the thinning distance. By default, it computes the Haversine distance using the `rdist.earth()` function from the **fields** package. If `distance = "euclidean"`, it computes the Euclidean distance instead. The main advantage of this method is that it computes the full distance matrix between your points. However, the main disadvantage is that it is very time and memory consuming, as it requires computing all pairwise comparisons.

```{r}
brute_thin <- thin_points(
  data = sim_data,
  method = "distance",
  search_type = "brute",
  thin_dist = 20,
  trials = 10,
  all_trials = FALSE,
  seed = 123
)
nrow(largest(brute_thin))
```

### Traditional *kd*-tree

A *kd*-tree is a binary tree that partitions space into regions to optimize nearest neighbor searches. The `search_type="kd_tree"` method uses the **nabor** R package to implement *kd*-trees via the **libnabo** library. This method is more memory-efficient than brute force. Since this implementation of *kd*-trees is based on Euclidean distance, when working with geographic coordinates (`distance = "haversine"`), **GeoThinneR** transforms the coordinates into XYZ Cartesian coordinates. Although this method scales better than the brute-force method for large datasets, since we want to identify all neighboring points of each query point, its performance can still be improved for very large datasets.

```{r}
kd_tree_thin <- thin_points(
  data = sim_data,
  method = "distance",
  search_type = "kd_tree",
  thin_dist = 20,
  trials = 10,
  all_trials = FALSE,
  seed = 123
)
nrow(largest(kd_tree_thin))
```

### Local *kd*-trees

To reduce the complexity of searching within a single huge *kd*-tree containing all points from the dataset, we propose a simple modification: creating small local *kd*-trees by dividing the spatial domain into multiple smaller regions, each with its own tree (`search_type="local_kd_tree"`). This notably reduces memory usage, allowing this thinning method to be used for very large datasets on personal computers. Additionally, you can run in parallel this method to improve speed using the `n_cores` parameter.

```{r}
local_kd_tree_thin <- thin_points(
  data = sim_data,
  method = "distance",
  search_type = "local_kd_tree",
  thin_dist = 20,
  trials = 10,
  all_trials = FALSE,
  seed = 123
)
nrow(largest(local_kd_tree_thin))
```

### Restricted neighbor search

Another way to reduce the computational complexity of *kd*-tree neighbor searches is by limiting the number of neighbors to find. In the `nabor::knn()` function, we can specify the number of neighbors to search for (`k`). For spatial thinning, we need to identify all neighbors for each point, which increases computational cost by setting `k` to the total number of points (`n`). However, with the `search_type="k_estimation"` method, we can estimate the maximum number of neighbors a point can have based on the spatial density of points, thereby reducing the `k` parameter. This method is particularly useful when datasets are not highly clustered, as it estimates a smaller `k` value, significantly reducing computational complexity.

```{r}
k_estimation_thin <- thin_points(
  data = sim_data,
  method = "distance",
  search_type = "k_estimation",
  thin_dist = 20,
  trials = 10,
  all_trials = FALSE,
  seed = 123
)
nrow(largest(k_estimation_thin))
```

## Grid-based thinning

Grid sampling is a standard method where the area is divided into a grid, and `n` points are sampled from each grid cell. This method is very fast and memory-efficient. The only drawback is that you cannot ensure a minimum distance between points. There are two main ways to apply grid sampling in **GeoThinneR**: (i) define the characteristics of the grid, or (ii) pass your own grid as a raster (`terra::SpatRaster`).

For the first method, you can use the `thin_dist` parameter to define the grid cell size (the distance in km will be approximated to degrees to define the grid cell size, one degree roughly 111.32 km), or you can pass the resolution of the grid (e.g., `resolution = 0.25` for 0.25x0.25-degree cells). If you want to align the grid with external data or covariate layers, you can pass the `origin` argument as a tuple of two values (e.g., `c(0, 0)`). Similarly, you can specify the coordinate reference system (CRS) of your grid (`crs`).

```{r}
system.time(
grid_thin <- thin_points(
  data = sim_data,
  method = "grid",
  resolution = 1,
  origin = c(0, 0),
  crs = "epsg:4326",
  n = 1, # one point per grid cell
  trials = 10,
  all_trials = FALSE,
  seed = 123
))
nrow(largest(grid_thin))
```

Alternatively, you can pass a `terra::SpatRaster` object, and that grid will be used for the thinning process.

```{r, message=FALSE, warning=FALSE}
rast_obj <- terra::rast(xmin = -10, xmax = 10, ymin = -5, ymax = 5, res = 1)
system.time(
grid_raster_thin <- thin_points(
  data = sim_data,
  method = "grid", 
  raster_obj = rast_obj,
  trials = 10,
  n = 1,
  all_trials = FALSE,
  seed = 123
))
nrow(largest(grid_raster_thin))
```
```{r, echo = FALSE}
rast_obj <- terra::as.polygons(rast_obj)
rast_obj <- sf::st_as_sf(rast_obj)
ggplot() +
  geom_sf(data = rast_obj, color = "#353839") +
  geom_point(data = sim_data, aes(x = lon, y = lat),color = "#EB714B", alpha = 0.5) +
  geom_point(data = largest(grid_raster_thin), aes(x = lon, y = lat),color = "#5183B3", alpha = 1) +
  labs(title = "Grid-based thinning results",
       x = "Longitude",
       y = "Latitude",
       caption = "Blue points are kept, red points are deleted") +
  theme_minimal()
```

## Precision thinning

In this approach, coordinates are rounded to a certain precision to remove points that fall too close together. After removing points based on coordinate precision, the coordinate values are restored to their original locations. This is the simplest method and is more of a filtering when working with data from different sources with varying coordinate precision. To use it, you need to define the `precision` parameter, indicating the number of decimals to which the coordinates should be rounded.

```{r}
system.time(
precision_thin <- thin_points(
  data = sim_data,
  method = "precision", 
  precision = 0,
  trials = 50,
  all_trials = FALSE,
  seed = 123
))
nrow(largest(precision_thin))
```
```{r, echo = FALSE}
ggplot() +
  geom_point(data = sim_data, aes(x = lon, y = lat),  color = "#EB714B", size = 3, alpha = 0.5) +
  geom_point(data = largest(precision_thin), aes(x = lon, y = lat), color = "#5183B3", size = 3) +
  labs(title = "Precision-based thinning results",
       x = "Longitude",
       y = "Latitude",
       caption = "Blue points are kept, red points are deleted") +
  theme_minimal()
```

These are the methods implemented in **GeoThinneR**. Depending on your specific dataset and research needs, one method may be more suitable than others.

# Additional customization features

## Spatial thinning by group

In some cases, your dataset may include different groups, such as species, time periods, areas, or conditions, that you want to thin independently. The `group_col` parameter allows you to specify the column containing the grouping factor, and the thinning will be performed separately for each group. For example, in the simulated data where we have two species, we can use this parameter to thin each species independently:

```{r}
all_thin <- thin_points(
  data = sim_data,
  thin_dist = 100,
  seed = 123
)
grouped_thin <- thin_points(
  data = sim_data,
  group_col = "sp",
  thin_dist = 100,
  seed = 123
)

nrow(largest(all_thin))
nrow(largest(grouped_thin))
```
```{r, echo =FALSE, fig.show="hold",  out.width="50%"}
removed <- sim_data[!all_thin$retained[[1]], ]
removed_group <- sim_data[!grouped_thin$retained[[1]], ]

ggplot() +
  geom_point(data = removed, aes(x = lon, y = lat, shape = sp), color = "#EB714B",  alpha = 0.2) +
  geom_point(data = largest(all_thin), aes(x = lon, y = lat, shape = sp), color = "#5183B3", alpha = 1) +
  scale_shape_manual(values = c(sp1 = 15, sp2 = 17)) +
  ggtitle("Thinning without grouping") +
  theme_minimal()

ggplot() +
  geom_point(data = removed_group, aes(x = lon, y = lat, shape = sp), color = "#EB714B",  alpha = 0.2) +
  geom_point(data = largest(grouped_thin), aes(x = lon, y = lat, shape = sp), color = "#5183B3", alpha = 1) +
  scale_shape_manual(values = c(sp1 = 15, sp2 = 17)) +
  ggtitle("Thinning by group") +
  theme_minimal()
```



## Fixed number of points

What if you need to retain a fixed number of points that best covers the area where your data points are located, or for class balancing? The `target_points` parameter allows you to specify the number of points to keep, and the function will return that number of points spaced as separated as possible. Additionally, you can also set a `thin_dist` parameter so that no points closer than this distance will be retained. Currently, this approach is only implemented using the brute force method (`method = "distance"` and `search_type = "brute"`), so be cautious when applying it to very large datasets.

```{r}
targeted_thin <- thin_points(
  data = caretta,
  lon_col = "decimalLongitude",
  lat_col = "decimalLatitude",
  target_points = 150,
  thin_dist = 30,
  all_trials = FALSE,
  seed = 123,
  verbose = TRUE
)
nrow(largest(targeted_thin))
```
```{r, echo = FALSE}
ggplot() +
  geom_sf(data = medit, color = "#353839", fill = "antiquewhite", alpha = 0.7)  +
  geom_point(data = largest(targeted_thin), aes(x = decimalLongitude, y = decimalLatitude),color = "#5183B3", alpha = 1) +
  xlab("Longitude") + ylab("Latitude") +
  ggtitle("Exact number of occurrences") +
  theme(
      panel.grid.major = element_line(color = gray(.5),linetype = "dashed",linewidth = 0.5),
      panel.background = element_rect(fill = "white"),
      axis.title = element_blank(),
      legend.position = "bottom"
    )
```

## Select points by priority

In some scenarios, you may want to prioritize certain points based on a specific variable, such as coordinate uncertainty, data quality, or any other variable. The `priority` parameter allows you to assign a numeric weight to each point, where higher values indicate higher importance, and these points are more likely to be retained during thinning. This feature is available for the `"distance"`, `"grid"` and `"precision"` methods.

For example, in the sea turtle data downloaded from GBIF, there is a column named `coordinateUncertaintyInMeters`. To prioritize records with lower uncertainty, you can invert the values so that smaller uncertainties receive higher priority. A simple way to do this is:

```{r}
# Invert uncertainty to give more priority to lower uncertainty records
priority <- -caretta$coordinateUncertaintyInMeters

# For example, handle NAs by assigning lowest possible priority
priority[is.na(priority)] <- min(priority, na.rm = TRUE) - 1
```

Then apply thinning using the priority weights:

```{r}
# Standard grid thinning (no priority)
grid_thin <- thin_points(
  data = caretta,
  lon_col = "decimalLongitude",
  lat_col = "decimalLatitude",
  method = "grid",
  resolution = 2,
  seed = 123
)

# Grid thinning with priority
priority_thin <- thin_points(
  data = caretta,
  lon_col = "decimalLongitude",
  lat_col = "decimalLatitude",
  method = "grid",
  resolution = 2,
  priority = priority,
  seed = 123
)

mean(largest(grid_thin)$coordinateUncertaintyInMeters, na.rm = TRUE)
mean(largest(priority_thin)$coordinateUncertaintyInMeters, na.rm = TRUE)
```

**Notes**

* Higher `priority` values mean the point is more likely to be retained.
* In `"grid"` and `"precision"` methods, when multiple points fall into the same grid cell or rounded coordinate, the one with highest priority is kept.
* In the `"distance"` method, priority is used to resolve tie situations where multiple points have the same spatial influence; the point with lowest priority is dropped.
* If multiple points have the same priority, one is selected at random.
* By default, if provided `NA` values in `priority`, they are treated as the lowest possible priority.

# Other Packages

The **GeoThinneR** package was inspired by the work of many others who have developed methods and packages for working with spatial data and thinning techniques. Our goal with **GeoThinneR** is to offer additional flexibility in method selection and to address specific needs we encountered while using other packages. We would like to mention other tools that may be suitable for your work:

- **spThin**: The `thin()` function provides a brute-force spatial thinning of data, maximizing the number of retained points through random iterative repetitions, using the Haversine distance.
- **enmSdmX**: Includes the `geoThin()` function, which calculates all pairwise distances between points for thinning purposes.
- **dismo**: The `gridSample()` function samples points using a grid as stratification, providing an efficient method for spatial thinning.

# References

- Aiello‐Lammens, M. E., Boria, R. A., Radosavljevic, A., Vilela, B., and Anderson, R. P. (2015). *spThin: an R package for spatial thinning of species occurrence records for use in ecological niche models*. Ecography, 38(5), 541-545.

- Elseberg, J., Magnenat, S., Siegwart, R., & Nüchter, A. (2012). *Comparison of nearest-neighbor-search strategies and implementations for efficient shape registration*. Journal of Software Engineering for Robotics, 3(1), 2-12.

- Hijmans, R.J., Phillips, S., Leathwick, J., & Elith, J. (2023). *dismo: Species Distribution Modeling*. R Package Version 1.3-14. https://cran.r-project.org/package=dismo

- Mestre-Tomás J (2025). GeoThinneR: Efficient Spatial Thinning of Species Occurrences. R package version 2.0.0, https://github.com/jmestret/GeoThinneR

- Smith, A. B., Murphy, S. J., Henderson, D., & Erickson, K. D. (2023). *Including imprecisely georeferenced specimens improves accuracy of species distribution models and estimates of niche breadth*. Global Ecology and Biogeography, 32(3), 342-355.

# Session info

```{r}
sessionInfo()
```
